---
title: BigArray - A DelayedArray backend for bigmemory
author:
- name: Malte Thodberg
  email: maltethodberg@gmail.com
output:
  BiocStyle::html_document:
    toc_float: yes
package: BigArray
vignette: >
  %\VignetteIndexEntry{User guide}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
```

# Introduction

The `BigArray` package implements a `r Biocpkg("DelayedArray")` backend for the
`r CRANpkg("bigmemory")` R-packages for handling massive out-of-memory
matrices. See https://petehaitch.github.io/BioC2020_DelayedArray_workshop/articles/Effectively_using_the_DelayedArray_framework_for_users.html for more information on `DelayedArray`.

# Installation

Install from github with devtools:

```{r eval=FALSE}
devtools::install_github("MalteThodberg/BigArray", 
                         dependencies = TRUE, 
                         build_vignettes = TRUE)
```

Load the package using:

```{r}
library(BigArray)
```

# Creating a `BigArray`

There a multiple ways to create a BigArray from and R matrix:

```{r}
X <- matrix(rnorm(1000), ncol=10)
```

Using the the `realization` machinery by coercing the matrix to a
`BigArray`, which will generate a temporary file backing file:

```{r}
as(X, "BigArray")
```

If you want to specify the location of the backing file you can use use
`writeBigArray`

```{r}
my_filematrix <- tempfile()
writeBigArray(X, filepath = my_filematrix)
```

Integer matrices are supported:

```{r}
Y <- round(X)
storage.mode(Y) <- "integer"
as(Y, "BigArray")
```

To use dimension names, you must first allow it via global settings from the bigmemory-package:

```{r}
# Use global settings to allow dim names
library(bigmemory)
options(bigmemory.allow.dimnames=TRUE)

# Set colnames and write
rownames(X) <- sprintf("GENE_%i", seq_len(nrow(X)))
colnames(X) <- sprintf("SAMP_%i", seq_len(ncol(X)))
writeBigArray(X)
```

Alternatively you can directly use functions from `bigmemory` and point
to the backing file afterwards.

```{r}
my_filematrix <- tempfile()

bm <-  filebacked.big.matrix(
    nrow = 100,
    ncol = 25,
    type = "double",
    backingfile = paste0(basename(my_filematrix), ".bmat"),
    backingpath = dirname(my_filematrix),
    descriptorfile = paste0(basename(my_filematrix), ".desc.txt"))

# Construct from bm
BigArray(my_filematrix)
```

Large .txt files can be processed similarly using
`bigmemory::read.big.matrix`.

# Manipulating `BigArray`s using delayed operations and DelayedMatrixStats

`BigArray`s are simply `DelayedArray` objects and can be manipulated as
such. The usual conventions for extracting data from matrix-like objects
work as expected:

```{r}
out <- as(X, "BigArray")
dim(out)
head(rownames(out))
head(out[,1])
```

We can also perform manipulations like subsetting and arithmetic. Note
that these operations do not affect the data in the bigmemory backend;
rather, they are delayed until the values are explicitly required, hence
the creation of the `DelayedMatrix` object.

```{r}
out[1:5,1:5] 
out * 2
out %*% runif(ncol(out))
```

We can also calculate matrix summary statistics using the `r Biocpkg("DelayedMatrixStats")` package, which supports block-wise processing:

```{r}
library(DelayedMatrixStats)
colSums2(out)
rowSums2(out)
rowVars(out)
```

# Using BigArray with SummarizedExperiment

TBA

# Session information

```{r}
sessionInfo()
```
